### 2016-10-13 documentation/paper.tex
Update documentation and LaTeX version

\documentclass[a4paper,11pt,twoside]{article}
\usepackage[utf8]{inputenc}
\usepackage{verbatim}
\usepackage[british]{babel}

\usepackage{url}

\usepackage{amsmath}
\usepackage{amsthm}

\usepackage[mathlf,fourierbb,minionint,opticals,fullfamily,swash]{MinionPro}
%\usepackage[mathlf,fourierbb,minionint]{MinionPro}
\usepackage{MnSymbol}
\usepackage{pifont}
%\usepackage[all]{xy}
\usepackage{graphicx}
%\usepackage[dvips]{graphicx}

\newcommand{\name}[1]{\textsf{#1}}

\usepackage[activate]{pdfcprot}
\usepackage[active]{srcltx}
\usepackage{fancyhdr}
\pagestyle{fancy}
%\fancyhead[RO,LE]{\thepage}
\fancyhead{}
\fancyhead[LE]{\nouppercase{\textssc{\leftmark}}}
\fancyhead[RO]{\nouppercase{\textssc{\rightmark}}}
\fancyfoot[C]{\thepage}
%\fancyfoot[C]{\Pisymbol{MinionPro-Extra}{120} \thepage\ \Pisymbol{MinionPro-Extra}{120}}
%\fancyfoot[LO]{\thepage}
%\fancyfoot[RE]{\thepage}
%\fancyfoot{}

\usepackage{enumitem}

\usepackage{listings}
\lstdefinestyle{customc}{
  belowcaptionskip=1\baselineskip,
  breaklines=true,
  xleftmargin=\parindent,
  showstringspaces=false,
  basicstyle=\small\ttfamily,
  columns=fullflexible,
  keepspaces=true,
}

\newcommand{\eq}{\ensuremath{\sim}}

\usepackage[explicit,toctitles]{titlesec}

\title{The Warrior Stabs the Goat: Creating a Dothraki Resource Grammar with Grammatical Framework}
\author{Benno Kuckuck}

\begin{document}

\begin{titlepage}
\clearpage{\pagestyle{empty}\cleardoublepage}
\phantom{bla}
%\vfill
\vfill
\begin{center}
\textsc{\Huge The Warrior Stabs the Goat}
\vspace{0.5cm}

{\LARGE Creating a Dothraki Resource Grammar with Grammatical Framework}

\vspace{2.5cm}

{\Large Maria Henkel}\\
2008709 $\bullet$ maria.henkel@hhu.de
\vspace{0.5cm}

{\Large Benno Kuckuck}\\
2340409 $\bullet$ benno.kuckuck@hhu.de
\vspace{1.5cm}

{\Large 17 October 2016}

\vspace{2.5cm}

\vfill
Lecturers:  Yulia Zinova, Rainer Osswald\\
Course: Introduction to ``Grammatical Framework'', SS 2016

\end{center}
\vfill
\end{titlepage}

\section{Introduction}
Grammatical Framework (GF) is a functional programming language, 
created in 1998 \cite{gf-book}, doubling as a categorial grammar 
formalism that can be used to implement (natural) language grammars. 
As GF is ``working from a language-independent representation of meaning''
\cite{wiki-gf}
it can be used to generate or parse text in different languages, 
e.g.\ for use in machine translation. 
Since GF uses a symbolic approach to processing and translating language,
in most cases its translations are more precise than those of the 
more commonly used statistical tools \cite{brown-frederking}. 
For example, on somewhat complicated input,
the ``Google Translate'' service \cite{google} will in many cases
produce ungrammatical sentences, while the GF translator 
\cite{gf-translate} has no problems producing 
a correct translation:

\begin{center}
  \begin{tabular}{p{3.7cm}p{4cm}p{3.7cm}}
    Input & Output Google Translate & Output GF Translator\\\hline
    I heal these women. & 
    Ich diese Frauen zu heilen. & 
    Ich verheile diese Frauen.\\
    The warrior will kiss that cheese behind the tree. & 
    Der Krieger wird Kuss, der K\''ase hinter dem Baum. & 
    Der Krieger wird jenen K\''ase hinter dem Baum k\''ussen.\\
    Diese hei\ss e Frau heilt diese K\''onigin im Meer. & 
    This hot woman heals the queen of the sea.  & 
    This hot woman heals this queen in the sea.
  \end{tabular}
\end{center}

Symbolic formalisms like GF -- unlike Google Translate -- depend 
on a large knowledge base of predefined grammar rules 
to achieve high-quality language processing: Its 
own standard library, the GF Resource Grammar Library, covers 
more than 38 different languages \cite{gf-status} -- although many of 
them are not yet complete -- and is still growing. 
In 2010, it already consisted of more than 500,000 
lines of code \cite{gf-book}. 
As Grammatical Framework is an open-source project \cite{ranta-rgl} 
and still under development, everyone is invited to use it 
or even help further develop it, for example by adding languages to 
the library, which is the aim of the project described in this paper. 
Our goal is to write a GF resource grammar for the Dothraki 
language and, eventually, be able to add it to the GF 
Resource Grammar Library. The work described in this paper is meant to 
lay a foundation for this goal (see Section \ref{sec-goals}).
Dothraki is a constructed language, developed by the language creator and writer 
David J. Peterson \cite{peterson-book}. The Dothraki 
are a fictional race of nomadic horse warriors in the ``A Song of Ice and 
Fire'' book series by George R. R. Martin \cite{martin}. 
As the series was adapted for television by the television network HBO, 
the language bits found in Martin's books had to be extended 
to a fully functional language \cite{peterson-web}: The Dothraki language. 
Today, it has a vocabulary of over 1,400 words 
\cite{vocab}. Examples of 
Dothraki language grammar will appear
in Section \ref{sec-impl}, but much more information can 
be found in the book ``Living Language Dothraki'' \cite{living-language}, 
on Peterson's blog \cite{peterson-web} and on the Tongues of Ice and 
Fire Wiki (2016) among many other sources.

This paper aims to serve as a short documentation and explanation of our 
project. It will cover the project aim and scope, as well as its 
structure and an exposition of our implementation of
Dothraki grammar rules in GF (Section \ref{sec-impl}). 
We will also discuss problems and limitations and briefly describe 
future plans before we summarize and evaluate the project in its 
current state and our experience while working on it (Section \ref{sec-discuss}).

\section{The Project}\label{sec-project}
As mentioned before, the eventual aim of this project is to create a 
resource grammar for the Dothraki language and add it 
to the GF Resource Grammar Library, so that other Dothraki fans 
and/or GF enthusiasts may use it. While there are currently a 
few Dothraki language learning apps and online translators 
available (see \cite{cognitus}, \cite{funtrans}), 
these are very limited, primarily relying on word-for-word 
translation unaware of any grammar rules. An open-source 
Dothraki resource grammar for GF, which already provides a wide 
range of tools and options for integrating with other projects, 
would undoubtedly be a significant addition to the available range 
of Dothraki language resources.

\subsection{About Resource Grammars and the Standard Library}\label{sec-rgl}

GF, like other symbolic approaches to machine translation and language 
processing, is well-suited for providing high-quality output in more 
narrowly focused applications (as opposed to statistical approaches, whose 
strengths lie more in providing acceptable output in a very wide range 
of scenarios). Most applications however share a large part of the rules 
necessary to parse or construct correct sentences, namely those that are 
concerned with the morphology and syntax of the involved language(s). 
While these could be implemented ad-hoc for every application, it is much 
more economic to provide language-specific but application-independent 
rules as a reusable library. This is the goal of GF's Resource Grammar 
Library.

The idea is to free application programmers from having to deal with 
``specialized linguistic expertise'' (Ranta, 2009). 
By providing a common interface across languages, it becomes 
considerably easier (in some cases even trivial) for the 
application programmer to add support for additional languages. 
Considerable work has been and is being done by the GF developers 
and resource grammar programmers to make this possible:

\begin{quote}
The idea of using a grammar as a software library is, to our knowledge,
new in GF. It has required a considerable effort in the design 
and implementation of the GF programming language: a type system, 
a module system, compilation techniques, and optimizations. [...] 
The  effort
made in this work is supplemented by a substantial effort in the 
library itself.
The code included in the library is more than twice in size compared to the
implementation of GF.\hfill\cite{ranta-rgl}
\end{quote} 
This division of labour benefits not just the application programmers,
however, but also the linguistics experts implementing the 
application-independent libraries:\\ Firstly, because a resource 
grammar, once written, can ``serve an unlimited number of applications'' 
\cite[p.\ 200]{gf-book}, but also because ``the existing library 
specification will help to identify the linguistic issues 
and avoid pitfalls'' by providing a well thought-out, 
language-independent interface.

\subsection{Project Goals and Structure}\label{sec-goals}
While in the long-term, our work is aimed towards a full 
implementation of a Dothraki resource grammar, this was not a 
realistic goal for this term paper, for two main reasons: 
First, Ranta estimates that an interval of three to six months 
of full-time work is needed for an experienced GF programmer to 
implement a new resource grammar \cite[p.\ 222]{gf-book}, which far 
exceeds the time frame for this term paper. And secondly, 
the Dothraki language itself is still a work-in-progress, with 
significant parts of the grammar either as yet undocumented, or 
possibly non-existent, which at present makes a full implementation 
of all the grammatical functions covered by GF's RGL arguably 
impossible (for more on this point, see the discussion in Section \ref{sec-discuss}).

In Ranta's book \cite[pp.\ 237--245]{gf-book} (see Figure \ref{fig-ita}), a mini resource 
grammar for Italian is implemented as an example.

\begin{figure}[h]\label{fig-ita}
\centering
  \includegraphics[width=10cm]{ita.png}
  \caption{Mini Resource Grammar Modules and Dependencies \cite[p.\ 209]{gf-book}}
\end{figure}

Our work lies somewhere between this example and a full resource grammar. 
Like in the Italian example, the abstract grammar specification 
our resource grammar is built on, is only a subset of that of the full 
RGL. Unlike the Italian example, it adheres much more closely to the
structure of the full RGL (see Figure \ref{fig-rgl}), employing the same file structure 
and API, in hopes of easing a later expansion to a full resource grammar.
It is also much more extensive than Ranta's example, covering a much wider
range of grammatical constructions.\footnote{As a quick comparison, the example 
from the book implements 12 grammatical categories, whereas our Dothraki 
implementation covers 27 of the 48 categories of the full RGL. In terms 
of API methods implemented, the comparison is even starker: Ranta's 
example grammar implements 19 API methods, our grammar 
implements almost 200.}

\begin{figure}[h]\label{fig-rgl}
\centering
  \includegraphics[width=13.5cm]{rgl.jpg}
  \caption{Module Structure of a Full Resource Grammar \cite{gf-slides}}
\end{figure}

Just like the GF repositories \cite{detrez-camilleri}, 
our full project is hosted on GitHub \cite{github} so that others may see, use and contribute to it.

The GF Resource Grammar Library consists essentially of three parts:
\begin{enumerate}
  \item The \emph{API} is the outward-facing part of the library.
  It contains those modules which will be used by application
  grammar programmers. It, in turn, is composed of two parts:
  \begin{itemize}
    \item A (large) language-independent
  part (contained in the \name{api} folder of the RGL), primarily
  the module \name{Syntax}, which (roughly) consist of the module
  \name{Cat}, defining the abstract categories
  of GF's langauge-independent representation of syntax
  and the module \name{Constructors}, which defines
  a wide array of functions for building up larger syntactical
  components (e.g.\ noun phrases, verb phrases, clauses) 
  from constituent parts.
  \item A (smaller) part specific to each language of the RGL, 
  defined in a module called \name{Paradigms\emph{XXX}},
  which contains functions for constructing instances of the
  basic lexical categories (e.g.\ nouns, verbs, adjectives).
  \end{itemize}

  \item The API functions are defined in terms of more basic
  operations defined in the \emph{abstract part} of the RGL,
  \footnote{For example the API contains 22 overloads for the function
  \name{mkCl}, used to construct clauses, but almost
  all of these are based on the single method 
  \name{predVP} from module \name{Sentence}.}
  which is contained in the folder \name{abstract}

  These are the operations which must be implemented for each 
  language.

  \item The \emph{concrete part} of the RGL, consists
  of concrete implementations of all abstract modules
  for each language that is part of the RGL. These are mostly contained
  in the folders carrying the name of the language.

  Some of the abstract operations and categories are implemented similarly
  for almost all languages. Such implementations are contained
  in the \name{common} folder.

  \item Additionally, but optionally, many languages
  also supply a large dictionary of words available in
  that language (allowing application grammar writers
  to use e.g.\ \verb+rise_N+ directly, instead of having to use
  \verb+mk_V ``rise'' ``rose'' ``risen''+ and risking getting
  the irregular forms wrong).
\end{enumerate}

As mentioned before, our project structure adheres closely
to this template (see Figure \ref{fig-folders}):

\begin{itemize}
  \item The \name{simpleapi} and \name{simpleabstract} folder contain
  the API part and abstract part respectively.

  Their contents are selectively copied over from the corresponding
  files in the \name{api} and \name{abstract} folders of the full RGL.

  This allows us to build a working resource grammar, without
  having to implement every single category and operation
  first. While our Dothraki resource grammar grew and continues
  to grow these are gradually expanded, until they converge
  towards the full RGL.

  \item The \name{simpleenglish} folder conatins those
  parts of the English resource grammar, which are relevant
  for our trimmed-down API.

  \item The \name{common} folder is taken straight from
  the full RGL.

  \item Finally, the \name{simpledothraki} folder contains
  our original contribution, the (somewhat
  trimmed-down) resource grammar for Dothraki.
\end{itemize}

There are three more folders, which are not part of the resource grammar:
\begin{itemize}
  \item The folder \name{example} contains an application grammar
  that uses our Dothraki resource grammar and the English resource
  grammar.

  \item The folder \name{documentation} contains this documentation.

  \item The folder \name{vocabulary} contains a Python script
  used for scraping the wiki source code of 
  the publically available vocabulary list \cite{vocab} and
  automatically generating a GF-compatible Dothraki dictionary
  (see Section \ref{sec-vocab}).
\end{itemize}

\begin{figure}[h]\label{fig-folders}
\begin{itemize}[itemsep=0.5pt]
\item simpleapi
\item simpleabstract
\item simpledothraki
\item simpleenglish
\item common

\vspace{0.2cm}
\item example

\vspace{0.2cm}
\item documentation 
\item vocabulary
\end{itemize}
\caption{The project's folder structure}
\end{figure}

\subsection{Implementation of Grammatical Categories in Dothraki}\label{sec-cats}
In this section we will go into some of the design decisions 
involved in the implementation of the Dothraki resource grammar. 
In order to get a high-level overview, we will first take a look 
at the linearizations of some of the categories defined in the 
API part of the abstract resource grammar library, showing how 
Dothraki grammar can be fit into the GF scheme.

\subsubsection{Nouns and Noun Phrases}
The relevant categories here are \name{N} (nouns), 
\name{CN} (common nouns), \name{NP} (noun phrases), \name{Pron} (pronouns) 
and \name{PN} (proper names), which are realized as 
follows:\footnote{Here, and in the following examples,
the code snippets do not necessarily appear exactly like
this in the actual source code, since usually some parts are 
defined in the \name{ResDot} 
module and some in \name{CatSimpleDot}, for reasons of 
reusability and implementation hiding. However, in all cases,
the actual implementations are equivalent to the listings
supplied here.}

\lstset{style=customc}
\lstMakeShortInline{§}

\begin{lstlisting}
N, CN = {s : Number => Case => Str ; a : Animacy } ;
NP, Pron = {s : Case => Str ; agr : Agr } ; 
PN = {s : Case => Str } ;
\end{lstlisting}

These implementations follow in a fairly straightforward fashion 
from the basic features of nouns in the Dothraki language: 
In Dothraki, every noun falls into one of two classes, 
\emph{animate} and \emph{inanimate} (which play a very similar 
role to grammatical gender in other languages). 
The language uses five cases: nominative, accusative, 
genitive, ablative and allative. Hence, we define:

\begin{lstlisting}
Animacy = Anim | Inanim ;
Case = Nom | Gen | Acc | All | Abl ;
\end{lstlisting}

Noun phrases, as well as pronouns, have a fixed number and person, 
which is stored in the §agr§ field of type §Agr§, simply defined as follows:

\begin{lstlisting}
Agr = Ag Person Number ;
\end{lstlisting}

There does not currently seem to be a need to store animacy 
in the agreement record. This is unlike English, where the 
§NP§s gender can affect the verb phrase 
(''The boy loves himself.'' but ``The girl loves herself.'') 
and thus needs to be stored in the §NP§. 
Finally, proper names are (so far as we can tell) 
always animate and third person singular, so none of that 
information needs to be stored in §PN§.

In order to turn a common noun into a noun phrase, 
we usually need a determiner. The relevant categories here are 
§Num§ (numerals), §Quant§ (quantifiers), §Predet§ (predeterminers) 
and §Det§ (determiners, roughly composed of optional predeterminers, 
a quantifier and a numeral). 
Numerals in the proper sense, as well as predeterminers, 
are not yet implemented in our resource grammar 
(there is a lack of good information on determiner structure in 
Dothraki), but since the §Num§ category is used in the 
GF Resource Grammar Library not just for representing 
numerals but also to determine the number of 
the determiner, we have a dummy implementation in 
our resource grammar:

\begin{lstlisting}
Num = {s: Str ; n : Number} ;
Quant = {s : QuForm => Case => Str } ;
Det = {s : Animacy => Case => Str ; n : Number ; s2 : Str } ;
\end{lstlisting}

The §s§ field of §Num§ is always empty in our implementation, 
this is necessary for technical reasons related to the implementation 
of GFs parser \cite{google-groups}.

Quantifiers in Dothraki (such as ``jin'' \eq\ ``this'' and ``rek'' \eq\ ``that'') are inflected for animacy, number and case, but the singular and plural inanimate forms are always identical, so ‘QuForm’ is defined as follows:

\begin{lstlisting}
QuForm = QAnim Number | QInanim ;
\end{lstlisting}

Unlike a §Quant§, a §Det§ has a fixed number.
An interesting feature of determiners in Dothraki is 
that they can appear either before the noun 
(''jinak adra'' = ``this turtle'') or after (such as determiners built 
from possessive pronouns: ``adra anni'' = ``my turtle''). 
Conceivably,\footnote{we cannot currently say for certain, 
due to lack of examples and documentation on 
Dothraki determiner structure} a determiner could even be split 
into two parts (''my two turtles'' $\stackrel{?}=$ ``akat adrasi anni''). 
In anticipation of this, §Det§ has two §Str§ components, §s§ and §s2§. 
Since possessive pronouns are the only examples of postposed 
determiners we know so far, and these happen not to 
inflect with animacy or case, the §s2§ part is simply a §Str§, whereas 
the §s§ part is a table of inflected forms.

\subsubsection{Verbs and Verb Phrases}\label{sec-verbs}

The relevant categories here are §V§ (intransitive verbs), 
§V2§ (transitive verbs with an §NP§ complement) 
and §VP§ (verb phrases), whose lincats are as follows:

\begin{lstlisting}
param VFormPN = Pers1 Number | Pers2 | Pers3 Number ;
param VForm = 
APast Polarity Number 
  | APresent Polarity VFormPN 
  | AFuture Polarity VFormPN 
  | ImpFormal Polarity 
  | ImpInformal Polarity ;

oper Verb : Type = {
  s : VForm => Str ; 
  inf : Str ; 
  part : Str
  } ;

V = Verb ;
V2 = Verb ** {objCase : Case} ;
VPSlash = Verb ** {objCase : Case ; subjpost : Str} ;
VP = Verb ** {compl : Str; subjpost : Str} ;
\end{lstlisting}

The basic structure §Verb§ contains the infinitive and participle 
as well as all conjugated forms of a verb. 
The somewhat convoluted setup of the §VForm§ parameter type is 
owed to the peculiarities of conjugation in Dothraki: 
First off, somewhat unusually,\footnote{from the Indo-European
viewpoint, that is, not for natural languages in general} in all 
tenses and moods, 
verb forms vary with the polarity of the sentence 
(``Me dothra\textbf{e}'' = ``He rides'', ``Me vos dothra\textbf{o}'' = ``He does not ride''). 
In the present and future tenses, verbs are inflected for 
person and number, but the second person singular and plural 
forms are always identical (hence the definition of §VFormPN§). In 
the past tense, verbs are inflected for 
number only, not for person (but unlike in the present and 
future tenses, second person singular and plural forms do differ). 
There are also formal and informal imperative forms.

Transitive verbs are basically the same as intransitive verbs, 
but they also assign a case to their §NP§ complement (mostly 
accusative, but other cases do appear), which is 
stored in the §objCase§ field.

A verb phrase essentially consists of a verb with a 
complement. There is also a field §subjpost§ which again encodes 
a peculiar feature of Dothraki grammar: 
A number of constructions, which, in other languages, 
are rendered by verbal auxiliaries (can, must) or verbs with §VP§ complements 
(try to), are realized in Dothraki 
via non-inflecting particles (e.g.\ ``laz'' \eq\ ``can'' 
and ``eth'' \eq\ ``must''), which do not otherwise affect 
the conjugation of the main verb (``Me dothrak'' = ``He rides'', 
``Me laz dothrak'' = ``He can ride'', 
``Me kis dothrak'' = ``He tries to ride'')\footnote{This could be compared
to the way negation is expressed in many Indo-European languages,
namely by a non-inflecting particle (``not'', ``nicht'', ``ne \dots\ pas'', etc.),
which is adposed to the verb in some manner, but does not otherwise
affect inflection.}. These have to 
be regarded as part of the verb phrase but, 
syntactically, are not preposed to the verb phrase but postposed 
to the subject of the clause (this makes a difference, e.g., in 
relative clauses, which have VSO word order instead of the 
standard SVO word order). The same is true of passive constructions, 
which are indicated by ``nem'' postposed to the subject 
(``Lajak dothrae'' = ``The warrior rides'', `` Hrazef nem dothrae'' = 
``The horse is ridden''; note that the verb still agrees in number and person
with the syntactical subject of the clause).

§VPSlash§ represents a verb phrase missing a complement 
and structurally kind of sits between §V2§ and §VP§. 
It is generated from a §V2§ (via §SlashV2 : V2 -> SlashVP§) 
and can then either be turned into a §VP§ by adding an object 
(via §ComplSlash : VPSlash -> NP -> VP§’) or into a §ClSlash§ (clause 
missing an object) by adding a subject (via 
§SlashVP : VPSlash -> NP -> ClSlash§), to be used, e.g., in a relative 
sentence (``the goat which the warrior stabbed''). As such it does 
not have a complement, like §VP§, and retains the §objCase§ from §V2§, 
but it does have the §subjpost§ field from §VP§ (so we can form, e.g., 
``the goat which the warrior can stab'' = ``dorvi fin vindee lajak laz'').

In order to implement verbal auxiliaries such as ``can'' and ``must'', 
which are represented as §VV§s in GF, we also have a 
dummy lincat for §VV§:

\begin{lstlisting}
VV = {s : Str} ;
\end{lstlisting}

Since ``can'', ``must'' etc.\ are simply uninflected particles 
in Dothraki, this is enough to represent those. 
This should of course not be considered a final design, 
since it is unlikely that all verbs with verb phrase 
complement can be realized like this in Dothraki. 
However, there is currently very little documentation available 
on how other §VV§ verbs are realized, so this dummy 
implementation will have to do for now.

\subsubsection{(Relative) Clauses and Sentences}

There are no big surprises in the lincats of §Cl§ and §QCl§ 
(clauses and question clauses, respectively) or §S§ and §QS§ 
(sentences and question sentences):

\begin{lstlisting}
S, QS = {s : Str} ;
Cl, QCl = {s : Tense => Anteriority => Polarity => Str} ;
\end{lstlisting}

The lincats of these categories are fairly canonical in 
many languages and in Dothraki they are basically the same as 
in English or German. The main differences between these 
languages come from whether the linearization also depends 
on some kind of §Order§ parameter, for example distinguishing 
indirect from direct questions or subordinate from 
main clauses. It might be necessary to add something 
like that to our lincat later, if more information about, e.g.,
indirect questions in Dothraki becomes available, 
but for now there is no indication that we 
need such an extra parameter.

Relative clauses are more interesting. 
The relevant categories here are §RP§ (relative pronoun), §ClSlash§ 
(a clause missing an object), §RCl§ (relative clauses) and §RS§
(relative sentence):

\begin{lstlisting}
RP = {s : QuForm => Case => Str } ;
RCl = {s : Tense => Anteriority => Polarity => Animacy => Number => Str } ;
ClSlash = {s : Tense => Anteriority => Polarity => Str; 
    subj: Str ; objCase : Case} ;
RS = {s : Animacy => Number => Str } ;
\end{lstlisting}

Relative pronouns in Dothraki inflect according to the 
animacy and number of the noun phrase they attach to, 
and the case that the relativized noun phrase would have 
been assigned in the embedded relative sentence (similar to German). 
Hence, the linearization of §RS§ depends on the 
animacy and number of the noun phrase 
it is eventually attached to. §RCl§ is to §RS§ 
what §Cl§ is to §S§, i.e., it is a relative sentence, 
whose tense, anteriority and polarity have not yet 
been determined. §ClSlash§ has to remember the case of 
the missing object, so it can assign it to the relative 
pronoun, if used in a relative sentence 
(via §RelSlash : RP -> ClSlash -> RCl§).

There is one more important difference between §Cl§ and §ClSlash§. 
When a clause is constructed from a verb phrase and 
subject via §PredVP : NP -> VP -> Cl§, the noun phrase is 
just prepended to the verb phrase. We cannot do this 
when constructing a §ClSlash§ via §SlashVP : NP -> VPSlash -> ClSlash§, 
because of a quirk in Dothraki word order: 
Whereas regular main clauses and questions 
follow SVO word order, relative clauses have 
VSO word order. So depending on whether a §ClSlash§ is eventually 
turned into a relative sentence (via §RelSlash : RP -> ClSlash -> RCl§) 
or a question (via §QuestSlash : IP -> ClSlash -> QCl§), we have 
to choose the word order accordingly 
(``dorvi fin vindee lajak'' = ``the goat which the warrior stabs'' vs 
``Fin lajak vindee?'' = ``What does the warrior stab?''). 
Therefore, §ClSlash§ contains the (linearized) subject in 
a field, §subj§, instead of embedding it in the §s§.

\subsubsection{Adjectives}
The relevant categories here are §A§ (adjectives), §AP§ 
(adjectival phrases) and §Comp§ (''complement of a copula'', 
though see below):

\begin{lstlisting}
A = {s : Degree => Number => ACase => Str ; 
    pred : VForm => Str };
AP = {s : Number => ACase => Str ; pred : VForm => Str } ;
Comp = {s : VForm => Str} ; 
\end{lstlisting}

There are two basic uses for adjectival phrases: 
attributive (as in ``the strong warrior'') and predicative 
(as in ``the warrior is strong''). Attributive use works 
similarly in Dothraki to, e.g., English and German: The 
adjectival phrase inflects for number and case in agreement 
with the noun it modifies. In Dothraki the forms for all 
the cases other than nominative are always identical, 
however, so we define

\begin{lstlisting}
ACase = ANom | AOther ;
\end{lstlisting}

instead of using §Case§ and having to duplicate these forms 
in the table.

In English or German, predicative use of an adjective 
employs a copula (``to be'' or ``sein''), which is inflected for 
tense, whereas the adjective is invariant (the same happens in 
sentences like ``the woman is a warrior'' or ``the man is on 
the mountain'', where the copula ``to be''
connects two noun phrases or a noun phrase and an adverbial phrase). 
Dothraki, however, is an 
entirely copulaless language. When an adjective is used 
predicatively, the adjective itself is essentially turned 
into a verb and inflected for tense and polarity:
\begin{quote}
``the strong warrior'' =
``lajak haj''\\
``The warrior is strong'' = ``lajak haja''\\
``the warrior will be strong'' = ``lajak vahaja''\\
``the warrior was not strong'' = ``lajak vos ahajo''
\end{quote}

Similarly, in a sentence like 
``the woman is a warrior'', the §NP§ itself inflects for tense,
with the ablative and allative form indicating past 
and future tense, respectively: 
\begin{quote}
``The woman is a warrior'' = ``Chiori lajak''\\
``The woman was a warrior'' = ``Chiori lajakoon''\\
``The woman will be a warrior'' = ``Chiori lajakaan''
\end{quote}

As a result, the category §Comp§ (i.e., ``strong'' in ``the warrior 
is strong'' and ``a warrior'' in ``the woman is a warrior'') essentially 
behaves like an intransitive verb in Dothraki. 
§A§ and §AP§ have to contain all conjugated forms 
of the verb as it might be used in a predicative context.

Indeed our lincats for §A§, §AP§ and §Comp§, while very much unlike 
the lincats in the English and German resource grammars, 
are quite similar to the lincats in the resource grammar 
for Japanese, another language in which the adjective 
itself can be inflected for tense and polarity in predicative 
use via a complex set of suffixes. As in our resource 
grammar, this leads to §AP§ and §Comp§ being structurally 
very similar to §VP§:
\begin{lstlisting}
VP = {
    verb : Speaker => Animateness => Style => TTense => Polarity => Str ;  
    a_stem, i_stem : Speaker => Animateness => Style => Str ;  
    te, ba : Speaker => Animateness => Style => Polarity => Str ;
    prep : Str ; obj : Style => Str ; 
    prepositive : Style => Str ; needSubject : Bool} ;
Comp = {
    verb : Animateness => Style => TTense => Polarity => Str ;
    a_stem, i_stem : Animateness => Style => Str ; 
    te, ba : Animateness => Style => Polarity => Str ;
    obj : Style => Str ; prepositive : Style => Str ; 
    needSubject : Bool} ;
AP = {
    pred : Style => TTense => Polarity => Str ; 
    attr, adv, dropNaEnging, prepositive : Style => Str ; 
    te, ba : Style => Polarity => Str ; 
    needSubject : Bool} ;
\end{lstlisting}

\subsubsection{Other categories}

Adverbial phrases, like in most other languages, are simply 
uninflected token strings, as implemented in §CommonX§:

\begin{lstlisting}
Adv = {s : Str} ; 
Prep = {s : Str ; c : Case} ;
\end{lstlisting}

Adverbial phrases can be formed from noun phrases 
using prepositions, which assign case in Dothraki, 
or sometimes are expressed \emph{only} by case. 
For example, §to_Prep§ from module §Structural§ is implemented 
in Dothraki simply as §{s = [] ; c = All}§: What is expressed in 
other laguages via the preposition ``to'', is understood in 
Dothraki from the modified noun phrase appearing in allative case.

Conjunctions in Dothraki have different forms 
depending on whether they are used to conjoin 
noun phrases/sentences, or as phrasal conjunctions to start a sentence:

\begin{lstlisting}
Conj = {s : Str; p : Str; n : Number} ;
\end{lstlisting}

E.g., the conjunction ``and'', when used to connect noun phrases,
is ``ma'' (as in ``ma lajak ma khaleesi'' = ``the warrior and the queen''), 
but translates as ``majin'', when used as a phrasal 
conjunction (''Majin lajak zoqwe khaleesies.'' = 
``And the warrior kissed the queen.'').

\begin{figure}\label{fig-cats}
  \centering
  \includegraphics[width=13cm]{cats.png}
  \caption{GF Resource Grammar Library Categories \cite{gf-slides}}
\end{figure}

Figure \ref{fig-cats} shows the main categories of the GF Resource 
Grammar Library and the current status of our implementation. 
The yellow categories have been fully or at least partially 
implemented in our resource grammar.

\subsection{Implementation of syntactical functions in Dothraki}

\subsubsection{Morphology}

Dothraki, generally speaking, has a much richer morphology than 
English, making the constructors for lexical categories 
supplied in §ParadigmsSimpleDot§ fairly complex. As an example, 
let us describe the construction of verbs 
(nouns and adjectives present similar challenges). 
Here is the constructor §mk2V : Str -> Str -> V§ 
which builds a §V§ from the infinitive form and the 
past singular (the two forms which would often be given in dictionaries):

\begin{lstlisting}
mk2V : Str -> Str -> Verb = \zalat,zal -> let {stem = stemV zalat zal} in {
  inf = zalat ;
  s = case stem of {
    fati@(fat + ("a"|"e"|"i"|"o")) => table {
      APast Pos Sg => zal ;
      APast Pos Pl => fati + "sh" ;
      APast Neg Sg => fat + "o" ;
      APast Neg Pl => fat + "osh" ;

      APresent pol pn => presForm fati pol pn ;

      AFuture Pos pn => presForm (pre {"a"|"e"|"i"|"o" => "v" ; _ => "a"} + fati) Pos pn ;
      AFuture Neg pn => presForm (pre {"a"|"e"|"i"|"o" => "v" ; _ => "o"} + fati) Neg pn ;

      ImpFormal Pos => stem ;
      ImpFormal Neg => fat + "o" ;

      ImpInformal Pos => fati + "s" ;
      ImpInformal Neg => fat + "os"
    } ;
    em => table {
      APast Pos Sg => zal ;
      APast Pos Pl => em + "ish" ;
      APast Neg Sg => em + "o" ;
      APast Neg Pl => em + "osh" ;

      APresent pol pn => presForm em pol pn ;

      AFuture Pos pn => presForm (pre {"a"|"e"|"i"|"o" => "v" ; _ => "a"} + em) Pos pn ;
      AFuture Neg pn => presForm (pre {"a"|"e"|"i"|"o" => "v" ; _ => "o"} + em) Neg pn ;

      ImpFormal Pos => em + "i" ;
      ImpFormal Neg => em + "o" ;

      ImpInformal Pos => em + "as" ;
      ImpInformal Neg => em + "os"
    } 
  } ;
  part = case stem of {
    fat + ("a"|"e"|"i"|"o") => stem + "y" ;
    _ => stem + "ay"
  }
} ;
\end{lstlisting}

Here §presForm : Str -> Polarity -> VFormPN -> Str§ is another 
function, that generates the present tense forms of a verb. 
This is reused for the future tense forms, which have 
the same endings, but are distinguished by a prefix.

\begin{lstlisting}
presForm : Str -> Polarity -> VFormPN -> Str = \stem,pol,pn -> case stem of {
  fati@(fat + ("a"|"e"|"i"|"o")) => case <pol,pn> of {
    <Pos, Pers1 Sg> => fati + "k" ;
    <Pos, Pers1 Pl> => fati + "ki" ;
    <Pos, Pers2> => fati + "e" ;
    <Pos, Pers3 Sg> => fati + "e" ;
    <Pos, Pers3 Pl> => fati + "e" ;

    <Neg, Pers1 Sg> => fat + "ok" ;
    <Neg, Pers1 Pl> => fat + "oki" ;
    <Neg, Pers2> => fati + "o" ;
    <Neg, Pers3 Sg> => fati + "o" ;
    <Neg, Pers3 Pl> => fati + "o"
  } ;
  em => case <pol,pn> of {
    <Pos, Pers1 Sg> => em + "ak" ;
    <Pos, Pers1 Pl> => em + "aki" ;
    <Pos, Pers2> => em + "i" ;
    <Pos, Pers3 Sg> => em + "a" ;
    <Pos, Pers3 Pl> => em + "i" ;

    <Neg, Pers1 Sg> => em + "ok" ;
    <Neg, Pers1 Pl> => em + "oki" ;
    <Neg, Pers2> => em + "i" ;
    <Neg, Pers3 Sg> => em + "o" ;
    <Neg, Pers3 Pl> => em + "i"
  }
} ;
\end{lstlisting}

As can be seen here, suffixes often differ according to 
whether the stem ends in a vowel or a consonant, which is 
handled by GF's pattern matching mechanisms. Future tense 
forms are indicated by a prefix (``v'' if the stem starts with a vowel, ``a''
if the stem starts with a consonant), which we implement employing 
GF's special ``prefix-dependent choice'' type.

In case the past singular form of the verb is not explicitly 
supplied, it needs to be guessed, which can be complicated:

\begin{lstlisting}
mk1V : Str -> Verb = \w -> case w of {
  ezo + "lat" => mk2V w ezo ;
  riss + "at" => mk2V w (addepenthesis riss) 
} ;
\end{lstlisting}

In general, the past singular form is just the stem of the verb, 
which mostly is derived by removing the "-lat" or "-at" 
infinitive ending.\footnote{Some verbs, such as "zalat" end in 
"-lat" but the "l" is part of the stem, 
not the infinitive ending. This cannot be guessed and has 
to be indicated in a dictionary.} However, this is complicated by the 
rules of epenthesis: The stem of "rissat" is "riss-", but the 
past singular is "risse", because a Dothraki word cannot end 
in "-ss". The rules for which consonant clusters may appear at 
the end of a word are complicated and for the most part captured in 
the function §addepenthesis : Str -> Str§, which adds an epenthesis 
"-e" to a string, if it ends in a disallowed consonant cluster:

\begin{lstlisting}
addepenthesis : Str -> Str = \w -> case w of {
  _ + ("a"|"e"|"i"|"o") => w ;
  _ + ("w"|"g"|"q") => w + "e" ;
  x + ("y"|"r"|"l") => case x of {
    _ + ("a"|"e"|"i"|"o") => w ;
    _ => w + "e"
  } ;
  x + ("m"|"n") => case x of {
    _ + ("a"|"e"|"i"|"o") => w ;
    y + ("w"|"y"|"r"|"l") => case y of {
      _ + ("a"|"e"|"i"|"o") => w ;
      _ => w + "e"
    } ;
    _ => w + "e"
  } ;
  x + ("th"|"s"|"sh"|"z"|"zh"|"kh"|"f"|"v") => case x of {
    _ + ("a"|"e"|"i"|"o") => w ;
    y + ("w"|"y"|"r"|"l"|"m"|"n") => case y of {
      _ + ("a"|"e"|"i"|"o") => w ;
      _ => w + "e"
    } ;
    _ => w + "e" 
  } ;
  x + ("j"|"ch"|"t"|"d"|"k") => case x of {
    _ + ("a"|"e"|"i"|"o") => w ;
    y + "w"|"y"|"r"|"l"|"m"|"n"|"th"|"s"|"sh"|"z"|"zh"|"kh"|"f"|"v") => case y of {
      _ + ("a"|"e"|"i"|"o") => w ;
      _ => w + "e"
    } ;
    _ => w + "e"
  }
} ;
\end{lstlisting}

This heuristic will give the right result in most (but not all) 
cases. Like in other resource grammars, there is always the 
option of supplying an irregular form by hand, if the heuristic fails.

\subsubsection{Syntax}

Many of the considerations about Dothraki syntax, 
that have gone into the design of our resource grammar, 
have already been mentioned in \ref{sec-cats}
As one example of how the design of our lincats is used to implement syntactical functions, let us go through the various ways of constructing clauses:

\begin{lstlisting}
PredVP np vp = {s = \\t,a,p =>
  np.s!Nom
  ++ vp.subjpost 
  ++ verbStr vp t a p np.agr
  ++ vp.compl ;
} ;
\end{lstlisting}

The function §PredVP : NP -> VP -> Cl§ from module 
§Sentence§ is the basic way of constructing a clause from a 
subject and a verb phrase (23 out of the 31 overloads 
of §mkCl§ from the resource grammar API are implemented in terms 
of §PredVP§). In our implementation this function concatenates 
four elements:

\begin{enumerate}
  \item The nominative form of the subject noun phrase,
  \item any particles indicating the mood of the verb phrase 
  (such as ``nem'' indicating passive voice, or ``laz''/``eth'' corresponding 
  roughly to ``can''/``must'' in English)
  \item the verb inflected for tense, anteriority, 
  polarity and person/number of the noun phrase,
  \item and the complement of the verb phrase.
\end{enumerate}

Conjugation of the verb is implemented in the function 
§verbStr§ defined in the module §ResDot§:
\begin{lstlisting}
verbStr : Verb -> Tense -> Anteriority -> Polarity -> Agr -> Str = \v,t,a,p,agr -> let vf = (tapaToVForm t a p agr) in 
  case <t,a> of {
    <Present,Anter> => "ray" ;    -- the perfect tense marker
    _ => []   
  } ++
  case p of {
    Pos => []  ;
    Neg => "vos" 
  } ++
  v.s!vf ;
\end{lstlisting}

The argument §vp§ in §PredVP§ above is of type §VP§ which is 
defined as §VP = Verb ** {compl : Str; subjpost : Str}§. 
GF has a mostly structural (as opposed to nominative) 
type system and thus §VP§ is automatically subtype of §Verb§, 
and variables of §VP§ type can be used where a §Verb§ is expected
(as we do above).

The function §verbStr§ is reused in other functions concerned 
with the construction of clauses.

\begin{lstlisting}
SlashVP np vpsl = {
  s = \\t,a,p => verbStr vpsl t a p np.agr ;
  subj = np.s!Nom ++ vpsl.subjpost;
  objCase = vpsl.objCase
} ;
\end{lstlisting}

The function §SlashVP : NP -> VPSlash -> ClSlash§ is used 
in several overrides of the API functions §mkClSlash§ as well as 
in some overrides of §mkQCl§ and §mkRCl§. Since §VPSlash§ is 
also an extension of §Verb§, we can reuse §verbStr§ to 
conjugate the verb. The subject phrase is not prepended 
to the verb (for reasons explained in Section \ref{sec-verbs}), but instead 
kept in the field §subj§. Unlike §VP§, a §VPSlash§ has no complement, 
but the case of the missing object has to be remembered.

\begin{lstlisting}
RelVP rp vp = {
  s = \\t,a,p,anim,n =>
    rp.s!(anToQuForm anim n)!Nom
    ++ vp.subjpost
    ++ verbStr vp t a p (Ag P3 n)
    ++ vp.compl ;
} ;
\end{lstlisting}

The function §RelVP : RP -> VP -> RCl§ is the basic way of 
creating relative clauses in which the subject of the sentence 
is being relativized (22 of the 25 overrides of §mkRCl§ are 
based on §RelVP§). This implementation is quite similar 
to §PredVP§, but with the relative pronoun taking the place 
of the subject, and being inflected for animacy and number.

The verb agrees in number with the relative pronoun and 
is always in the third person.

Finally, clauses can be formed using §ExistNP : NP -> Cl§.

\begin{lstlisting}
ExistNP np = {s = \\t,a,p => 
  np.s!Gen
  ++ (verbStr vekhat_V t a p np.agr)
} ;
\end{lstlisting}

This is for creating sentences such as ``Lajaki vekha'' = ``There is a warrior''. 
Unlike any other verb in Dothraki, the verb ``vekhat'' assigns 
genitive case 
to the subject of the sentence. Apart from that, 
we again use §verbStr§ to conjugate the verb ``vekhat'' 
(§vekhat_V§ is defined in module §ExtraDot§).

\subsection{Voca

With the current implementation status, it is already possible 
to parse and translate a wide range of phrases. The following 
example text is fully supported by the current version of the 
Dothraki resource grammar:

Beispieltext…

3.  Discussion 
3.1 Limitations
Warum nicht alles auf einmal? Zu viel Arbeit, alles gleichzeitig, keine Möglichkeit zu testen, ….

For the implementation of a resource grammar, a good grammar source is needed (Grammatical Framework, 2016b). Although the full resource grammar for the Dothraki language is not yet complete, it is becoming harder and harder to find reliable resources regarding further aspects of grammar we would like to implement (e.g. usage of structural words). As it is a constructed language, and relatively new, we hope that further and more precise information will be documented and published with time. 
Yet, even if we had all the information needed for a full implementation, it would not have been possible for us to write a full resource grammar in such a short amount of time, especially as we are still new to the GF programming language and not experienced when it comes to implementing resource grammars. Hence, we focus on providing a good basis of implemented rules, so that we – and others – can expand the project in the future.

Incompleteness/overgeneration of GF approach (grammar rules vs. natural language)….
Regeln versuchen so viel wie mögl. abzudecken, aber nicht immer möglich, siehe Buch? Zitat?...
3.2 Outlook
Expand grammar to a full resource grammar. Community?
3.3 Conclusion
Sinnvolles projekt, ambitioniertes ziel, viel gelernt, nicht einfach, stolz…

Appendix
Source Code
The complete source code can be viewed and downloaded via GitHub: 
\url{https://github.com/mahen2/dothraki_gf}

\makeatletter
\renewcommand\@biblabel[1]{}
\renewenvironment{thebibliography}[1]
     {\section*{\refname}%
      \@mkboth{\MakeUppercase\refname}{\MakeUppercase\refname}%
      \list{}%
           {\leftmargin0pt
            \@openbib@code
            \usecounter{enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\makeatother

\begin{thebibliography}{9}
\bibitem[Brown \& Frederking, 1995]{brown-frederking}
Brown, R.\ \& Frederking, R.\ (1995). 
\emph{Applying statistical English language modeling to symbolic machine 
translation}. 
In Proceedings of the 
sixth international conference on theoretical and methodological 
issues in machine translation (TMI-95), pp. 221--239.

\bibitem[Cognitus Apps, 2016]{cognitus}
Cognitus Apps (2016). 
Flamingo Dothraki. 
Retrieved from \url{https://play.google.com/store/apps/details?id=com.cognitusapps.flamingo.dothraki&hl=de}

\bibitem[Fun Translations, 2016]{funtrans}
Fun Translations (2016). Dothraki translator. 
Retrieved from \texttt{http://funtranslations.com/dothraki}

\bibitem[D\'etrez\ \& Camilleri, 2016]{detrez-camilleri}
D\'etrez, G.\ \&\ Camilleri, J.J.\ (2016). \emph{Grammatical Framework}. 
GitHub.com. Retrieved from \texttt{https://github.com/GrammaticalFramework}

\bibitem[Google, 2016]{google}
Google (2016). \emph{Google Translate}. Retrieved from 
\url{https://translate.google.com/}

\bibitem[Google Groups, 2016]{google-groups}
Google Groups (2016). \emph{Problem with strange metavariables in parsing}.
Retrieved from \url{https://groups.google.com/d/msg/gf-dev/aRjt_2JfvA0/iTcZbTu5AAAJ)}

\bibitem[Grammatical Framework, 2016a]{gf-translate}
Grammatical Framework (2016a). 
\emph{Translation with GF: Powered by multilingual grammars}. 
Retrieved from \url{http://www.grammaticalframework.org/demos/translation.html}

\bibitem[Grammatical Framework, 2016b]{gf-rgl}
Grammatical Framework (2016b). 
\emph{GF Resource Grammar Library: Synopsis}. 
Retrieved from \url{http://www.grammaticalframework.org/lib/doc/synopsis.html}

\bibitem[Martin, 2016]{martin}
Martin, G.R.R. (2016). \emph{Bibliography}. 
Retrieved from \url{http://www.georgerrmartin.com/bibliography/}

\bibitem[Henkel \& Kuckuck, 2016]{github}
Henkel, M.\ \& Kuckuck, B.\ (2016). 
\emph{Building a resource grammar for Dothraki}. GitHub.com. 
Retrieved from \url{https://github.com/mahen2/dothraki_gf}

\bibitem[Peterson, 2014]{living-language}
Peterson, D.J.\ (2014). 
\emph{Living Language Dothraki: A Conversational Language Course 
Based on the Hit Original HBO Series Game of Thrones.}
New York: Living Language.

\bibitem[Peterson, 2015]{peterson-book}
Peterson, D.J.\ (2015). \emph{The Art of Language Invention.}
Penguin Books.

\bibitem[Peterson, 2016]{peterson-web}
Peterson, D.J.\ (2016). \emph{About Dothraki}. 
Retrieved from \url{http://www.dothraki.com/about-dothraki/}

\bibitem[Ranta, 2009]{ranta-rgl}
Ranta, A.\ (2009). 
\emph{The GF Resource Grammar Library.}
Linguistic Issues in Language Technology, 2(2), pp.\ 1--63.

\bibitem[Ranta, 2011a]{gf-book}
Ranta, A.\ (2011a). 
\emph{Grammatical Framework: Programming with Multilingual Grammars.}
Stanford: CSLI Publications.

\bibitem[Ranta, 2011b]{gf-slides}
Ranta, A.\ (2011b). 
\emph{Grammatical Framework: Programming with Multilingual Grammars. 
Slides for the GF book}. Retrieved from 
\url{http://www.grammaticalframework.org/gf-book/gf-book-slides.pdf}

\bibitem[Ranta, 2015]{gf-status}
Ranta, A.\ (2015). 
\emph{The Status of the GF Resource Grammar Library}. Grammatical Framework. 
Retrieved from \url{http://www.grammaticalframework.org/lib/doc/status.html}

\bibitem[Tongues of Ice and Fire Wiki, 2016]{vocab}
Tongues of Ice and Fire Wiki (2016). \emph{Vocabulary}. 
Retrieved from \url{http://wiki.dothraki.org/Vocabulary}

\bibitem[Wikipedia, 2016]{wiki-gf}
Wikipedia (2016). \emph{Grammatical Framework}.
Retrieved from \url{https://en.wikipedia.org/wiki/Grammatical_Framework}
\end{thebibliography}

\end{document}

### 2016-10-14 documentation/paper.tex
Update documentation

@@
 \documentclass[a4paper,11pt,twoside]{article}
+
+\makeatletter
+
+\newcommand\frontmatter{%
+    \cleardoublepage
+  %\@mainmatterfalse
+  \pagenumbering{roman}}
+
+\newcommand\mainmatter{%
+    \cleardoublepage
+ % \@mainmattertrue
+  \pagenumbering{arabic}}
+
+\newcommand\backmatter{%
+  \if@openright
+    \cleardoublepage
+  \else
+    \clearpage
+  \fi
+ % \@mainmatterfalse
+   }
+
+\makeatother
+
 \usepackage[utf8]{inputenc}
@@
   belowcaptionskip=1\baselineskip,
+  escapeinside=``,
   breaklines=true,
@@
 
+\usepackage[table]{xcolor}
+
+\hyphenation{Doth-ra-ki}
+
 \title{The Warrior Stabs the Goat: Creating a Dothraki Resource Grammar with Grammatical Framework}
@@
 \begin{document}
-
+\frontmatter
 \begin{titlepage}
@@
 %\vfill
-\vfill
 \begin{center}
+\includegraphics[width=5cm]{hhu-logo.jpg}
+
+\vspace{2cm}
+
+{\Large Term Paper}
+\vspace{1cm}
+\hrule
+\vspace{0.4cm}
 \textsc{\Huge The Warrior Stabs the Goat}
@@
 
+\vspace{0.4cm}
+\hrule
 \vspace{2.5cm}
@@
 
+\tableofcontents
+
+\newpage
+
+\thispagestyle{empty}
+\begin{center}
+\Large
+  \includegraphics[width=5cm]{drogo.png}
+  \vspace{1.5cm}
+
+  \begin{minipage}{10cm}
+  \centering
+  ``Khal Drogo shillo chiories she krazaaj. Chiori fin vijazer 
+  adraes zoqwe Khales Drogo. Khal Drogo vinde ma chiories ma 
+  adraes she krazaaj. Khal Drogo vachrara.''
+
+  \vspace{0.7cm}
+  \hrule
+  \vspace{0.3cm}
+
+  \color{gray}
+``Khal Drogo met a woman on a mountain. The woman, who 
+protected a turtle, kissed Khal Drogo. Khal Drogo stabbed the 
+woman and the turtle on the mountain. Khal Drogo stinks.''
+
+\vspace{0.5cm}
+\color{black}
+GF Run-Time System, 2016
+  \end{minipage}
+\end{center}
+
+\newpage
+
+\thispagestyle{empty}
+
+\mainmatter
 \section{Introduction}
@@
 \begin{center}
-  \begin{tabular}{p{3.7cm}p{4cm}p{3.7cm}}
+  \rowcolors{2}{gray!25}{white}
+  \begin{tabular}{p{3.7cm}|p{4cm}|p{3.7cm}}
+  \rowcolor{gray!50}
     Input & Output Google Translate & Output GF Translator\\\hline
@@
     The warrior will kiss that cheese behind the tree. & 
-    Der Krieger wird Kuss, der K\''ase hinter dem Baum. & 
-    Der Krieger wird jenen K\''ase hinter dem Baum k\''ussen.\\
-    Diese hei\ss e Frau heilt diese K\''onigin im Meer. & 
+    Der Krieger wird Kuss, der K\"ase hinter dem Baum. & 
+    Der Krieger wird jenen K\"ase hinter dem Baum k\"ussen.\\
+    Diese hei\ss e Frau heilt diese K\"onigin im Meer. & 
     This hot woman heals the queen of the sea.  & 
@@
 
-\begin{figure}[h]\label{fig-ita}
+\begin{figure}[h]
 \centering
   \includegraphics[width=10cm]{ita.png}
-  \caption{Mini Resource Grammar Modules and Dependencies \cite[p.\ 209]{gf-book}}
+  \caption{\label{fig-ita}Mini Resource Grammar Modules and Dependencies \cite[p.\ 209]{gf-book}}
 \end{figure}
@@
 
-\begin{figure}[h]\label{fig-rgl}
+\begin{figure}[h]
 \centering
   \includegraphics[width=13.5cm]{rgl.jpg}
-  \caption{Module Structure of a Full Resource Grammar \cite{gf-slides}}
+  \caption{\label{fig-rgl}Module Structure of a Full Resource Grammar \cite{gf-slides}}
 \end{figure}
@@
   to use e.g.\ \verb+rise_N+ directly, instead of having to use
-  \verb+mk_V ``rise'' ``rose'' ``risen''+ and risking getting
+  \verb+mk_V "rise" "rose" "risen"+ and risking getting
   the irregular forms wrong).
@@
 
-\begin{figure}[h]\label{fig-folders}
+\begin{figure}[h]
 \begin{itemize}[itemsep=0.5pt]
@@
 \end{itemize}
-\caption{The project's folder structure}
+\caption{\label{fig-folders}The project's folder structure}
 \end{figure}
 
-\subsection{Implementation of Grammatical Categories in Dothraki}\label{sec-cats}
+\subsection{Implementation of Grammatical Categories in Dothraki}
+\label{sec-impl}
 In this section we will go into some of the design decisions 
@@
 §NP§s gender can affect the verb phrase 
-(''The boy loves himself.'' but ``The girl loves herself.'') 
+(``The boy loves himself.'' but ``The girl loves herself.'') 
 and thus needs to be stored in the §NP§. 
@@
 that they can appear either before the noun 
-(''jinak adra'' = ``this turtle'') or after (such as determiners built 
+(``jinak adra'' = ``this turtle'') or after (such as determiners built 
 from possessive pronouns: ``adra anni'' = ``my turtle''). 
@@
 Dothraki determiner structure} a determiner could even be split 
-into two parts (''my two turtles'' $\stackrel{?}=$ ``akat adrasi anni''). 
+into two parts (``my two turtles'' $\stackrel{?}=$ ``akat adrasi anni''). 
 In anticipation of this, §Det§ has two §Str§ components, §s§ and §s2§. 
@@
 
-\subsubsection{Other categories}
+\subsubsection{Other Categories}
 
@@
 
-\begin{figure}\label{fig-cats}
+\begin{figure}
   \centering
   \includegraphics[width=13cm]{cats.png}
-  \caption{GF Resource Grammar Library Categories \cite{gf-slides}}
+  \caption{\label{fig-cats}GF Resource Grammar Library Categories \cite{gf-slides}}
 \end{figure}
@@
 
-\subsection{Implementation of syntactical functions in Dothraki}
+\subsection{Implementation of Syntactical Functions in Dothraki}
 
@@
 that have gone into the design of our resource grammar, 
-have already been mentioned in \ref{sec-cats}
-As one example of how the design of our lincats is used to implement syntactical functions, let us go through the various ways of constructing clauses:
+have already been mentioned in \ref{sec-impl}.
+As one example of how the design of our lincats is used to 
+implement syntactical functions, let us go through the 
+various ways of constructing clauses:
 
@@
 
-\subsection{Voca
+\subsection{Vocabulary Extraction}\label{sec-vocab}
+
+\subsubsection{Dictionaries and Lexicons in the RGL}
+As mentioned earlier, many of the resource grammars in GF's
+library also come with a dictionary often called \name{Dict\emph{XXX}}.
+
+For example, in \name{DictGer} from the German resource grammar, we
+find entries such as:
+\begin{lstlisting}
+entziehen_V = irregV "entziehen" "entzieht" "entzog" "entz`\"o`ge" 
+  "entzogen" ; 
+kreis_N = mkN  "Kreis" "Kreise" masculine ;
+\end{lstlisting}
+An application grammar writer can then use these definitions and does not
+have to worry about the gender of nouns, plural forms 
+or irregular conjugations.
+
+In fact, the dictionaries in the RGL come in the
+form of an abstract grammar with a concrete implementation. The
+abstract grammar, often called \name{Dict\emph{XXX}Abs}, contains
+entries like
+\begin{lstlisting}
+entziehen_V : V ;
+kreis_N : N ;
+\end{lstlisting}
+
+These dictionaries, of course, cannot be used directly for translation.
+However, there are two ways in which the RGL supports direct
+translation between two langauges. First, there is the abstract
+grammar \name{Lexicon} in the \name{abstract} folder, which covers
+about 350 more or less commonly used words:
+\begin{lstlisting}
+...
+drink_V2 : V2 ;
+dry_A : A ;
+dull_A : A ;
+dust_N : N ;
+...
+\end{lstlisting}
+This abstract grammar is then implemented in many of the languages
+of the RGL. From \name{LexiconGer}:\footnote{The choice of names here is unfortunate:
+In the common linguistic sense of the words, \name{DictGer} is a lexicon, while
+\name{LexiconGer} is a dictionary...}
+\begin{lstlisting}
+drink_V2 = dirV2 Irreg.trinken_V ;
+dry_A = regA "trocken" ;
+dull_A = regA "stumpf" ;
+dust_N = reg2N "Staub" "St`\"a`ube" masculine ;
+\end{lstlisting}
+
+The \name{Lexicon} module is part of the abstract grammar
+\name{Lang} and its concrete implementations
+in various languages. Of course, the vocabulary is far
+too small to be useful for applications, but it is useful for
+experimentation and for testing the resource grammars.
+
+The RGL also has a folder \name{translator} which
+mirrors this structure on a much larger scale:
+The abstract grammar \name{Dictionary} contains 65,000 English 
+words,\footnote{as well as 450 Finnish proper names...} and
+the concrete implementations \name{Dictionary\emph{XXX}} contain
+their translations into various languages. This is used, e.g.,
+in the GF online translator \cite{gf-translate}.
+
+\subsubsection{The Dothraki dictionary}
+
+The publicly available vocabulary list for Dothraki \cite{vocab} currently
+contains about 1,500 words.\footnote{This is still not enough to implement
+the abstract \name{Lexicon} module from GF's resource grammar library:
+Many common words do not yet have a Dothraki equivalent.}
+In order to make these accessible for GF application grammars
+and for the purpose of translation from and to English, we
+supply, as part of our project, a Python script that translates
+the wiki source of this list into a GF-compatible format.
+
+The script creates five files in total, adhering closely 
+to the structure established in the full RGL, as explained above:
+\begin{itemize}
+  \item the abstract grammar \name{DictDotAbs.gf} and 
+  its concrete implementation \name{DictDot.gf} in the
+  folder \name{simpledothraki},
+  \item the abstract grammar \name{BigLexicon.gf} in
+  \name{simpleabstract} and its implementations
+  \name{BigLexiconEng.gf} and \name{BigLexiconDot.gf} in
+  \name{simpleenglish} and \name{simpledothraki}, respectively.
+\end{itemize}
+
+\begin{figure}
+\begin{tabular}{l|l}
+\begin{minipage}{.44\linewidth}
+\begin{lstlisting}
+*'''aranikh''' [a?a'nix]
+:''ni.'' subject, topic, theme
+
+*'''gerat''' [ge'?at]
+:''vtr.&rarr;abl.'' to lack
+:{{Vsup|ger}}
+\end{lstlisting}
+\end{minipage}&
+\begin{minipage}{.55\linewidth}
+\begin{lstlisting}
+`\textsf{in simpledothraki/DictDotAbs.gf}:`
+aranikh_N : N ;
+gerat_V2 : V2 ;
+
+`\textsf{in simpledothraki/DictDot.gf}:`
+aranikh_N = mkN "aranikh" inanimate ;
+gerat_V2 = mkV2 "gerat" "ger" ablative ;
+
+`\textsf{in simpleabstract/BigLexicon.gf}:`
+subject_N : N ;
+topic_N : N ;
+theme_N : N ;
+lack_V2 : V2 ;
+
+`\textsf{in simpledothraki/BigLexiconDot.gf}:`
+subject_N = DictSimpleEng.subject_N;
+topic_N = DictSimpleEng.topic_N;
+theme_N = DictSimpleEng.theme_N;
+lack_V2 = DictDot.gerat_V2;
+
+`\textsf{in simpleenglish/BigLexiconEng.gf}:`
+subject_N = DictDot.aranikh_N;
+topic_N = DictDot.aranikh_N;
+theme_N = DictDot.aranikh_N;
+lack_V2 = DictSimpleEng.lack_V2;
+\end{lstlisting}
+\end{minipage}
+\end{tabular}
+\caption{\label{fig-vocab}Two vocabulary wiki entries and generated GF definitions}
+\end{figure}
+
+Figure \ref{fig-vocab} shows an example of the GF code the script
+generates. A word is added to \name{BigLexicon} whenever a
+corresponding word is found in \name{DictEng}, the massive
+English dictionary that is part of the RGL.
+
+The wiki vocabulary list has 1,380 entries at the time
+of this writing. From these, our script creates
+1,410 definitions in \name{DictDot}.\footnote{the somewhat larger
+number comes from the fact that many words can be used
+in two different ways, for example both as an intransitive
+and as a transitive verb, giving more than one entry in \name{DictDot}}
+
+There are a total of 1,916 English translations in the vocabulary list.
+The script is able to extract 1,059 entries for \name{BigLexicon}
+from this. Many translations cannot be directly added to
+\name{BigLexicon} for various reasons:
+\begin{itemize}
+  \item Some English words appear as translations for
+  more than one Dothraki word. The script picks the first
+  one it finds.\footnote{Obviously, that might not result in the
+  best or most appropriate translation being picked, but that issue
+  can be addressed by later manual editing of the generated list.}
+  \item Many Dothraki words do not correspond to any one English word.
+  For example the list contains entries for ``defensive sword strike'',
+  ``sound of fire going out'', ``roast quail'' or
+  ``time period of approximately two minutes''. While for some of these,
+  it might be desirable to have dictionary 
+  entries,\footnote{For example, there could be an entry
+  §roast_quail_CN : CN§ in \name{BigLexicon} with linearizations
+  §roast_quail_CN = mkCN DictEng.roast_A (mkCN DictEng.quail_N)§
+  in
+  \name{BigLexiconEng} and §roast_quail_CN = mkCN DictDot.mechikh_N§
+  in \name{BigLexiconDot}.} this process is almost impossible
+  to automate and must be done by hand.
+\end{itemize}
+
+While \name{DictDot} and \name{DictDotAbs} are meant to be used
+in application grammars, the \name{BigLexicon\emph{XXX}} files
+can be used for testing and translation. An example of this
+is shown in the next section.
+
+\subsection{Example}
 
 With the current implementation status, it is already possible 
-to parse and translate a wide range of phrases. The following 
-example text is fully supported by the current version of the 
-Dothraki resource grammar:
-
-Beispieltext…
-
-3.  Discussion 
-3.1 Limitations
-Warum nicht alles auf einmal? Zu viel Arbeit, alles gleichzeitig, keine Möglichkeit zu testen, ….
-
-For the implementation of a resource grammar, a good grammar source is needed (Grammatical Framework, 2016b).
-Although the full resource grammar for the Dothraki language is not yet complete, it is becoming harder and harder to find reliable resources regarding further aspects of grammar we would like to implement (e.g.
-usage of structural words).
-As it is a constructed language, and relatively new, we hope that further and more precise information will be documented and published with time.
-Yet, even if we had all the information needed for a full implementation, it would not have been possible for us to write a full resource grammar in such a short amount of time, especially as we are still new to the GF programming language and not experienced when it comes to implementing resource grammars.
-Hence, we focus on providing a good basis of implemented rules, so that we – and others – can expand the project in the future.
-
-Incompleteness/overgeneration of GF approach (grammar rules vs. natural language)….
-Regeln versuchen so viel wie mögl. abzudecken, aber nicht immer möglich, siehe Buch? Zitat?...
-3.2 Outlook
-Expand grammar to a full resource grammar. Community?
-3.3 Conclusion
-Sinnvolles projekt, ambitioniertes ziel, viel gelernt, nicht einfach, stolz…
-
-Appendix
-Source Code
+to parse and translate a wide range of phrases. For experimentation
+and testing we supply the \name{LangSimple\emph{XXX}} abstract and
+conrete grammars which simply import the \name{SyntaxSimple\emph{XXX}}
+and \name{BigLexicon\emph{XXX}} modules (completely
+analogous to the \name{Lang\emph{XXX}} module in the
+full RGL) and can be used directly
+from the GF command line:
+
+\begin{lstlisting}
+> i simpleenglish/LangSimpleEng.gf
+linking ... OK
+
+Languages: LangSimpleEng
+2745 msec
+LangSimple> i simpledothraki/LangSimpleDot.gf
+linking ... OK
+
+Languages: LangSimpleDot LangSimpleEng
+187 msec
+LangSimple> p "the warrior which protected the turtle has stabbed my goat with a dagger ." | l
+lajak fin viljazer adraes ray vinde dorv anni ma mihesofoon .
+the warrior which protected the turtle has stabbed my goat with a dagger .
+lajak fin viljazer adraes ray vinde dorv ma mihesofoon anni .
+the warrior which protected the turtle has stabbed my goat with a dagger .
+
+31 msec
+\end{lstlisting}
+
+The multiple outputs for the last command are 
+quite typical and derive from the fact that the parse of the English 
+sentence is ambiguous:
+Was the duck stabbed with a dagger or was a duck with a dagger stabbed?
+Note that, in this case, the Dothraki translations differ, because the
+possessive pronoun ``anni'' follows the noun phrase
+(either ``dorv'' = ``goat'' or ``dorv ma mihesofoon'' = ``goat with a dagger'').
+
+\section{Discussion}\label{sec-discuss}
+
+To conclude our project documentation, we will now discuss problems 
+and limitations as well as briefly describe future plans before 
+we summarize the project in its current state and our experience 
+while working on it.
+
+\subsection{Limitations}
+To begin programming a resource grammar for the first time was 
+not easy, since it is not possible to start with, for example, 
+just one category and then test the implementation. As there are 
+so many dependencies in the resource grammar library, it is recommended 
+to take the resource grammar of a similar language and change it step 
+by step. But this way, as long as our implementation of the Dothraki 
+language was incomplete, the resource grammar would fill the gaps with 
+the wrong language grammar. Another way is to start out with an 
+independent mini resource grammar, but in that case you cannot make 
+use of the resources of the library. Finding our own solution for
+this problem took time, but in the end we are satisfied with how
+it turned out.
+
+For the implementation of a resource grammar, a good grammar
+source is needed (Grammatical Framework, 2016b). Although the
+full resource grammar for the Dothraki language is not yet complete,
+it is becoming harder and harder to find reliable resources regarding
+further aspects of grammar we would like to implement (e.g.
+usage of structural words). As it is a constructed language, and
+relatively new, we hope that further and more precise information
+will be documented and published with time. 
+Yet, even if we had all the information needed for a full implementation,
+it would not have been possible for us to write a full resource grammar
+in such a short amount of time, especially as we are still new to the
+GF programming language and not experienced when it comes to implementing
+resource grammars. Hence, we focus on providing a good basis of implemented
+rules, so that we – and others – can expand the project in the future.
+
+\subsection{Outlook}
+In the future, we would like to expand our resource grammar to a
+full resource grammar and add it to the GF resource grammar library.
+We are already working on and hosting the project online, for everyone
+to see, so that the GF and/or Dothraki community may help us to achieve
+this goal. Some more parts we can implement right now, for some we have
+to wait until more information regarding the Dothraki language is being
+published. In the end, we also want to add the full vocabulary to our
+dictionary.
+
+\subsection{Conclusion}
+Now that the first milestone of our project lies behind us,
+we can evaluate the working process and the result. As we had
+some starting problems at the beginning of the project, we are
+proud to have been able to create such a big portion of the
+resource grammar in the given time. By relying on the basics
+learned during the seminar and by aiming for such an ambitious
+goal, we surely learned a lot during the process. As we are
+convinced that our project can be helpful and interesting for
+other people in the future, we were motivated to keep on working
+on the resource grammar and to try to find the best solution for
+every problem. This term paper does not mark the end of this project,
+but we are curious to hear what you, our lecturers, and other people
+think of the work we have done so far.
+
+\backmatter
+
+\section*{Appendix}
+
+\paragraph{Source Code}
+
 The complete source code can be viewed and downloaded via GitHub: 

### 2016-10-14 documentation/paper.tex
More discussion in doc

@@
 
+\subsection{Difficulties}
+Beginning to write a resource grammar for the first time was 
+not easy. Due to the structure of the full RGL, it is not 
+possible to implement a resource grammar for a new language
+step-by-step, adding linearizations for a few important
+categories and expanding gradually, at least not when basing
+it directly on the abstract specifications from the full RGL.
+
+Of course, it is not feasible either to implement the entire
+resource grammar before being able to do any kind of testing.
+As a remedy for this problem, Ranta recommends
+taking the resource grammar of a similar language as a
+starting point and changing it step 
+by step \cite[p.\ 224]{gf-book}. 
+This proved difficult in our case, since Dothraki, as a constructed
+language, of course does not have family ties to any existing language. 
+While its grammar is certainly inspired by phenomena seen in
+natural languages, these appear to be picked from a variety
+of different language families (many of which, as non-linguists, we are not
+particularly familiar with). Furthermore, taking
+a language like English as a starting point, which is certainly similar
+to Dothraki in many aspects, but wildly different in others,
+proved to be fraught with difficulties as well. This is because
+of the web of dependencies in a grammar,
+changes at one point, such as increasing the number of cases 
+from two to five, tend to require changes in a number
+of different places all over the grammar, an effect that
+can quickly snowball. Finding our own solution for
+this problem, namely growing the abstract part gradually 
+(by selectively copying over parts from the full RGL) along with
+the concrete implementation, took time, but in the end we 
+are satisfied with how it turned out.
+
 \subsection{Limitations}
-To begin programming a resource grammar for the first time was 
-not easy, since it is not possible to start with, for example, 
-just one category and then test the implementation. As there are 
-so many dependencies in the resource grammar library, it is recommended 
-to take the resource grammar of a similar language and change it step 
-by step. But this way, as long as our implementation of the Dothraki 
-language was incomplete, the resource grammar would fill the gaps with 
-the wrong language grammar. Another way is to start out with an 
-independent mini resource grammar, but in that case you cannot make 
-use of the resources of the library. Finding our own solution for
-this problem took time, but in the end we are satisfied with how
-it turned out.
-
-For the implementation of a resource grammar, a good grammar
-source is needed (Grammatical Framework, 2016b). Although the
-full resource grammar for the Dothraki language is not yet complete,
-it is becoming harder and harder to find reliable resources regarding
-further aspects of grammar we would like to implement (e.g.
-usage of structural words). As it is a constructed language, and
-relatively new, we hope that further and more precise information
-will be documented and published with time. 
-Yet, even if we had all the information needed for a full implementation,
-it would not have been possible for us to write a full resource grammar
-in such a short amount of time, especially as we are still new to the
-GF programming language and not experienced when it comes to implementing
-resource grammars. Hence, we focus on providing a good basis of implemented
-rules, so that we – and others – can expand the project in the future.
+As Ranta notes in \cite{gf-slides}, the most important
+sources for creating a resource grammar are
+``[a] \emph{good} grammar book'' and ``[a] \emph{good} dictionary''.
+
+That poses some problems in the case of Dothraki.
+There are basically only two sources of information
+on Dothraki grammar: The writings of inventor 
+David J.\ Peterson \cite{peterson-web},
+the community-operated ``Tongues of Ice and Fire`` wiki
+at \url{wiki.dothraki.org}. These tend to be great resources
+for the basic parts of the language. Peterson's
+blog contains detailed, accessible explanations of
+many grammatical phenomena featured in
+the snippets of Dothraki, that have appeared on the show
+up to its third season. And the enthusiastic Dothraki community
+has done great work organizing and detailing this information.
+
+However, as our work progressed, it has become harder and harder
+to find good information on some of the more subtle or
+obscure parts of the grammar. It should be noted that the
+total corpus of text in the Dothraki language 
+still comprises only a few hundred sentences.
+By now, we have already mostly
+incorporated the information available from the two sources
+mentioned above (the two largest remaining omissions are wh-questions and
+numerals). In many cases we have run up against the limits
+of what information is available and sometimes we have been
+forced to resort to the Dothraki corpus itself
+in an effort to try and reverse engineer
+language features that so far have not been explained by
+their inventor.
+
+We give three examples:
+\begin{itemize}
+  \item As noted in Section \ref{sec-verbs} our resource grammar
+  so far only contains a dummy implementation of
+  verbs with a verb phrase complement (§VV§s). In English, 
+  (auxiliary) verbs like ``can'', ``must'' and ``want to'' are
+  of this kind. In Dothraki, as explained in the quoted section,
+  ``can'' and ``must'' are rendered as particles
+  (``laz'' and ``eth'') and this is what our current
+  implementation captures. However, certainly not all
+  §VV§s are rendered like this
+  in Dothraki. These non-inflected particles almost surely
+  are a small closed class of words, whereas §VV§s are
+  plentiful (think ``agree to (doing sth)'', ``consider (doing sth)'',
+  ``enjoy (doing sth)'', ``expect to (do sth)'' etc.).
+  However, there is almost no explanation available
+  on the syntactical structure of such verbs in Dothraki.
+
+  When they have appeared on the show, verb phrase
+  complements are usually avoided in favour of
+  embedded clauses:
+
+  \begin{quote}
+  ``The khaleesi wants to eat something different tonight.'' =
+  ``Khaleesi zala meme adakha esinakh ajjalan.'' (lit.\ ``The khaleesi wants that she eats something different tonight'')\\
+  ``You'd rather be sold into slavery?'' = ``Yer zali meyer nem vazhi ven zafra?'' (lit.\ ``You want that you are sold into slavery?'')
+  \end{quote}
+
+  As far as we can tell, in the entire corpus there are only four sentences
+  in which the verb ``zalat'' (to want) is used with what seems to be a
+  true verb phrase complement:
+  \begin{quote}
+    ``Hash shafka zali addrivat mae [...]?'' = ``Do you want me to kill him?''\\
+    ``[...] hash yer zali zifichelat moon?'' = ``[...] you want to steal from him?''\\
+    ``Anha zalak nesat ven fini ven athyazhar khaleesisi.'' = ``I'd like to know what a Khaleesi tastes like.''\\
+    ``Hash yeri vo zali nesat rek dirgak anha?'' = ``Don't you want to know what I think?''
+  \end{quote}
+  We could not find any other instances of §VV§ verbs being used
+  with a true verb phrase complement in the corpus.
+
+  From these four sentences it certainly appears like verb phrase complements
+  (similar to English) are formed with the verb in the infinitive
+  (``addrivat'', ``zifichelat'' and ``nesat'' are all infinitive forms).
+  However, as far as we can tell, this has not been discussed by 
+  Peterson so far, nor is there any information about it on the wiki.
+
+  It seems rather premature and speculative
+  to conclude from just four sentences (all featuring
+  the same verb) that this is the general way verb phrase
+  complements are rendered in Dothraki. Therefore, we are at an
+  impasse: Until further information becomes available,
+  we cannot implement §VV§s (other than the few
+  special ones like ``can'' and ``must'').
+
+  \item The module \name{Structural} contains a number
+  of common structural words that should be available in 
+  every language. Quite a few of these are simply not yet
+  available in Dothraki (``somewhere'', ``almost'',
+  ``quite'', ``otherwise'', ``at least'' etc.), but
+  even those that have translations in the vocabulary list
+  are sometimes difficult to implement. For example, the
+  vocabulary list notes that ``someone'' translates as
+  ``ato''. But since this is a noun phrase, it probably
+  must inflect somehow.
+  Inflection in nouns depends on their animacy, but
+  it is not clear whether inanimate or animate declension
+  is used here (or some irregular inflection, as with determiners). 
+  In the corpus ``ato'' appears exactly once,
+  in the nominative. In the end, we had no other choice
+  but to guess the declension (we decided on standard
+  animate inflection).
+
+  Similarly, ``eyak'' supposedly means ``everybody'', according
+  to the vocabulary, but in the only place it is used
+  in the corpus, it seems to be used in the sense of ``everything''.
+  In that instance it is used in the genitive (probably; the phrase
+  is ``hatif eyaki'' and the preposition ``hatif'' can assign
+  genitive, allative or ablative case, depending on meaning).
+  In the end we decided on a declension based on this one occurence and
+  by analogy with ``vosak'' (= ``nobody''), which occurs in
+  the accusative once.
+
+  In the end, in many cases in the \name{Structural} module, 
+  we had to employ something more akin to educated guesswork rather
+  than being able to rely on definitive sources.
+
+  \item §Comp§s (``complementizers to a copula'') can be used
+  to form imperatives: ``Be strong!'' or ``Be a warrior!''.
+  The first has an obvious rendering in Dothraki (by putting
+  the verb ``hajat'' = ``to be strong'' derived from ``haj'' = ``strong''
+  in the imperative), but the second does not. In Dothraki
+  there is no analogue of the copula ``to be'' that could be
+  used in the imperative here. In the corpus, no sentence
+  like this exists. For now, we cannot implement this
+  correctly.
+\end{itemize}
+
+We are hopeful that, in spite of these current limitations,
+our work so far can serve as a good basis for implementing
+more language features as further and more precise information 
+on them becomes available.
 

### 2016-10-16 documentation/paper.tex
Updated docs / some grammar fixes
I added some additional vocabulary and implemented mass nouns
in order to make the example text in the doc a bit nicer.

@@
 
-\usepackage[activate]{pdfcprot}
+%\usepackage[activate]{pdfcprot}
+\usepackage{microtype}
 \usepackage[active]{srcltx}
@@
 \usepackage[table]{xcolor}
+
+\usepackage{breakcites}
 
@@
 the ``Google Translate'' service \cite{google} will in many cases
-produce ungrammatical sentences, while the GF translator 
+produce ungrammatical or wrong translations, while the GF translator 
 \cite{gf-translate} has no problems producing 
-a correct translation:
+a correct translation (as of 22 September 2016):
 
@@
     Input & Output Google Translate & Output GF Translator\\\hline
+    Ich komme im Krieg um. & I'm in the war. & I die in the war.\\
+    Ich denke in der Vorlesung immer nach, bevor ich rede. &
+    I always think about the lecture before I speak. &
+    I always think in the lecture, before I talk.\\
     I heal these women. & 
@@
 on a large knowledge base of predefined grammar rules 
-to achieve high-quality language processing: Its 
+to achieve high-quality language processing:\footnote{As can be seen
+when trying to translate grammatical constructions that are not 
+explicitly covered by GF's grammar rules.
+For a taste of this, try translating ``Die Krone der K\"onigin ist sch\"on''
+or ``Das Buch, das die Frau mir gab, war alt'' to English using the
+GF translator (possessive constructions and object
+relativization for verbs taking more than one object are currently
+not covered satisfactorily in GF's abstract grammar model).}
+Its 
 own standard library, the GF Resource Grammar Library, covers 
@@
 however, but also the linguistics experts implementing the 
-application-independent libraries:\\ Firstly, because a resource 
+application-independent libraries: Firstly, because a resource 
 grammar, once written, can ``serve an unlimited number of applications'' 
@@
 \begin{figure}[h]
+\centering
+\fbox{
+\begin{minipage}{5cm}
 \begin{itemize}[itemsep=0.5pt]
@@
 \end{itemize}
-\caption{\label{fig-folders}The project's folder structure}
+\end{minipage}
+}
+\caption{\label{fig-folders}The project's folders}
 \end{figure}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{tabular}{l|l}
@@
 There are a total of 1,916 English translations in the vocabulary list.
-The script is able to extract 1,059 entries for \name{BigLexicon}
+The script is able to extract 1,130 entries for \name{BigLexicon}
 from this. Many translations cannot be directly added to
@@
 
+As an example of how extra vocabulary, that cannot be automatically
+extracted from the vocabulary list, can be supplied manually the module 
+\name{ExtraLex\emph{XXX}} contains a (currently minimal) sampling of
+additional words.
+
 While \name{DictDot} and \name{DictDotAbs} are meant to be used
@@
 possessive pronoun ``anni'' follows the noun phrase
-(either ``dorv'' = ``goat'' or ``dorv ma mihesofoon'' = ``goat with a dagger'').
+(either ``dorv'' = ``goat'' or ``dorv ma mihesofoon'' = ``goat with a 
+dagger'').\footnote{There are often \emph{lots} of different parses when
+parsing Dothraki sentences. Dothraki has no articles,
+does not distinguish between he/she/it and rarely distinguishes
+singular and plural for inanimate nouns. The sentence
+``Me vindee dorv'' can mean ``He stabs the goat'', ``She stabs the goat'',
+``He stabs goats'', etc. The three possible choices for ``me'' and
+five possible choices for ``dorv'' (singular/plural definite/indefinite article  or mass noun) 
+already give fifteen parses and that's
+not even counting all the \emph{really} strange interpretations
+deriving from the fact that ``vindee'' could also be the future 
+tense of ``indelat'' = ``to drink''...}
+
+Here is a more elaborate example:\footnote{The sentences are lightly editted compared
+to the parser input/output, to account for the
+fact that the lexer requires spaces around
+punctuation and first words of sentences may not be capitalized.
+There are also usually multiple possible parses, in which case we picked
+the most appropriate one.}
+
+\begin{quote}
+I found this woman on the mountain, Khal Drogo. 
+She was asleep and there was a dead lizard behind her. 
+Its head had been cut off. 
+She is a sorceress! 
+They squeeze the blood from the lizard and they drink it. 
+Then they crush the bones and they cook them with many roots and the fresh skins of snakes. 
+It is known. 
+She will try to poison you! 
+Do you believe me? 
+Therefore you must kill her! 
+Don't listen to her. 
+A woman who cooks poison mustn't survive. 
+She is dangerous. 
+Everybody knows. 
+You are the leader whom the horde respects. 
+Cut her throat with your arakh! 
+\end{quote}
+
+\begin{quote}
+Anha ez jinakes chiories she krazaaj, zhey Khal Drogo.
+Me remek ma zhavi driva irge mae vekh.
+Nhare mae nem ray zirisse.
+Mori efi qoy ha zhavoon ma mori indee mae.
+Hash mori kaffi tolor ma mori jolini mora ma ma sanoon garfothoon ma ilekoon choshi gezrisi.
+Me nem nesa.
+Me kis vizza shafka!
+Hash shafka shilloe anna?
+Majin shafka eth addrivi mae!
+Vos charos maan.
+Chiori fin jolina iz eth vos thiroo.
+Me zhowaka.
+Eyak nesa.
+Shafka akkelenak finnaan choma khalasar.
+Rissas foth mae ma arakhoon shafki!
+\end{quote}
+
+It demonstrates the following
+features supported by our resource grammar:
+\begin{itemize}[itemsep=0pt]
+  \item Transitive and intransitive verbs
+  \item Tenses: Past perfect, present, future, present perfect, past perfect
+  \item Negation
+  \item Passive voice
+  \item Verbal auxiliaries
+  \item Imperatives
+  \item Personal pronouns
+  \item Demonstratives
+  \item Possessive pronouns
+  \item Existential clauses
+  \item Adverbial phrases with prepositions
+  \item Zero-copula sentences
+  \item Adjectives (attributive and predicative)
+  \item Relative clauses (both subject and object relativization)
+  \item Proper names
+  \item Mass nouns
+  \item Vocatives
+  \item Yes/no questions
+  \item Conjunctions (for noun and verb phrases)
+  \item Phrasal conjunctions
+  \item Structural words (every, many, somebody, nobody, ...)
+\end{itemize}
 
@@
 
-\subsection{Limitations}
+\subsection{Limitations}\label{sec-lim}
 As Ranta notes in \cite{gf-slides}, the most important
@@
 on Dothraki grammar: The writings of inventor 
-David J.\ Peterson \cite{peterson-web},
+David J.\ Peterson \cite{peterson-web} and
 the community-operated ``Tongues of Ice and Fire`` wiki
-at \url{wiki.dothraki.org}. These tend to be great resources
+at \url{wiki.dothraki.org}. To be sure, these tend to be great resources
 for the basic parts of the language. Peterson's
@@
 many grammatical phenomena featured in
-the snippets of Dothraki, that have appeared on the show
+the snippets of Dothraki that have appeared on the show
 up to its third season. And the enthusiastic Dothraki community
-has done great work organizing and detailing this information.
+has done tremendous work in organizing and detailing this information.
 
@@
 incorporated the information available from the two sources
-mentioned above (the two largest remaining omissions are wh-questions and
-numerals). In many cases we have run up against the limits
+mentioned above. In many cases we have run up against the limits
 of what information is available and sometimes we have been
@@
 
-  When they have appeared on the show, verb phrase
+  Where §VV§s do appear in the corpus, verb phrase
   complements are usually avoided in favour of
@@
 
+  The problem is only more pronounced for the even more
+  obscure categories of verbs, such as §VA§, §VQ§, §V2A§, §V2Q§,
+  or §V2S§.\footnote{To give an idea: Of the around 65,000 words in \name{DictEng},
+  there are 49 of type §VA§, 7 of type §VQ§, 15 of type §V2A§, four of type
+  §V2S§ and one of type §V2Q§. Obviously, evidence of usage of such words in the
+  Dothraki corpus is scarce.}
+
   \item The module \name{Structural} contains a number
@@
   ``ato''. But since this is a noun phrase, it probably
-  must inflect somehow.
+  must inflect for case somehow.
   Inflection in nouns depends on their animacy, but
@@
   is used here (or some irregular inflection, as with determiners). 
-  In the corpus ``ato'' appears exactly once,
-  in the nominative. In the end, we had no other choice
+  In the corpus, ``ato'' appears exactly once,
+  in the nominative, thus giving no useful hints. In the end, we had no other choice
   but to guess the declension (we decided on standard
@@
   Similarly, ``eyak'' supposedly means ``everybody'', according
-  to the vocabulary, but in the only place it is used
-  in the corpus, it seems to be used in the sense of ``everything''.
+  to the vocabulary, but no declension is given and the word appears
+  only once in the corpus.\footnote{and curiously,
+  it seems to be used more in the sense of ``everything''}
   In that instance it is used in the genitive (probably; the phrase
@@
 \subsection{Outlook}
-In the future, we would like to expand our resource grammar to a
-full resource grammar and add it to the GF resource grammar library.
+The obvious goal for the future of our project is to expand the
+resource grammar further, hopefully to a point where it can eventually be added
+to the GF resource grammar library.
 We are already working on and hosting the project online, for everyone
 to see, so that the GF and/or Dothraki community may help us to achieve
-this goal. Some more parts we can implement right now, for some we have
+this goal. Some parts could probably be implemented right now 
+(the largest omissions right now being wh-questions and numerals; 
+§V3§s and reflexive verbs would be useful, too, but their implementation 
+would likely
+run up against some of the problems laid out in Section \ref{sec-lim}), 
+for some we have
 to wait until more information regarding the Dothraki language is being
-published. In the end, we also want to add the full vocabulary to our
-dictionary.
+published. Finally, it would also be useful to improve the dictionary
+by augmenting our current automatic extraction approach with
+some manual curation.
 
@@
 Now that the first milestone of our project lies behind us,
-we can evaluate the working process and the result. As we had
-some starting problems at the beginning of the project, we are
-proud to have been able to create such a big portion of the
+we can evaluate the working process and the result. After having
+overcome our initial problems setting up the project, we are
+proud of having been able to create such a big portion of the
 resource grammar in the given time. By relying on the basics
@@
 goal, we surely learned a lot during the process. As we are
-convinced that our project can be helpful and interesting for
+convinced that our project can be useful and interesting for
 other people in the future, we were motivated to keep on working
@@
 
-\backmatter
-
 \section*{Appendix}
@@
 \url{https://github.com/mahen2/dothraki_gf}
+
+\backmatter
 

### 2016-10-16 documentation/paper.tex
Typos

@@
 
-Of course, it is not feasible either to implement the entire
-resource grammar before being able to do any kind of testing.
+Of course, implementing the entire
+resource grammar before being able to do any kind of testing
+is not a feasible strategy either.
 As a remedy for this problem, Ranta recommends
@@
 
+\newpage
+\thispagestyle{empty}
+
+{\LARGE Versicherung}
+
+\vspace{1cm}
+Hiermit versichern wir, diese Arbeit selbst\"andig verfasst und keine anderen als
+die angegebenen Quellen benutzt zu haben. Wir versichern ferner, dass diese
+Arbeit oder Teile davon nicht schon für eine andere Abschlusspr\"ufung oder einen 
+anderen Beteiligungsnachweis eingereicht wurden. Uns ist bewusst, dass ein 
+Zuwiderhandeln gegen diese Versicherung eine Ordnungswidrigkeit darstellt, die 
+mit einer Geldbu\ss e geahndet werden kann.
+
+\vspace{1cm}
+D\"usseldorf, den 16.10.2016
+
 \end{document}

### 2016-10-17 vocabulary/extract.py
Update to extraction script

@@ -2,13 +2,13 @@
 
 # read filter from DictEng.gf
 
-filter = []
-with open("DictEng.gf", "r") as input_filter:
+filter = set()
+with open("../simpleenglish/DictSimpleEng.gf", "r") as input_filter:
     for line in input_filter.readlines():
         # only get the definitions
         if line.startswith('lin '):
             # only get the words
-            filter.append(line.split("=")[0].replace(
+            filter.add(line.split("=")[0].replace(
                 "lin ", "").replace("\n", "").strip())
 
 # read input
@@ -18,46 +18,61 @@ with open("input_dot.txt", "r", encoding="utf-8") as input_dot:
 
 data = data.split('\n*')
 
-output_org = []
-output_eng = []
-output_dict = []
-data = data[1:]
+output_dictabs = ["abstract DictDotAbs = CatSimple ** {","fun"]
+output_dictdot = ["concrete DictDot of DictDotAbs = CatSimpleDot ** open ParadigmsSimpleDot in {","lin"]
+output_lexabs = ["abstract BigLexicon = CatSimple ** {","fun"]
+output_lexeng = ["concrete BigLexiconEng of BigLexicon = CatSimpleEng ** open DictSimpleEng in {","lin"]
+output_lexdot = ["concrete BigLexiconDot of BigLexicon = CatSimpleDot ** open DictDot in {","lin"]
+
+# data = data[1:]
 type_lexicon = [
     ('adj.', 'A'),
     ('ni.', 'N'),
     ('na.', 'N'),
     ('v.', 'V'),  # beides: V, V2
-    ('vtr.', 'V2'),  # beides: V2, V3
+    ('vtr', 'V2'),  # beides: V2, V3
     ('vin.', 'V')]
 
+dictabs_str = '{wordident}_{w_type} : {w_type};'
+dictdot_str = '{wordident}_{w_type} = mk{w_type} "{word}" {gram_info};'
+lexabs_str = '{trans}_{w_type} : {w_type};'
+lexeng_str = '{trans}_{w_type} = DictSimpleEng.{trans}_{w_type};'
+lexdot_str = '{trans}_{w_type} = DictDot.{word}_{w_type};'
 
-def output_prep(w_type, gram_info, word, anim, trans):
+dictabs = set()
+
+def output_prep(w_type, gram_info, word, trans):
+    wordid = word + "_" + w_type
+    if wordid in dictabs:
+        return
+    dictabs.add(wordid)
[…]
